From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-26 Pi/paper/finite.tex
one pass addressing things from review A
Explain how the relation is lifted to presentation
Fixes #142

Co-authored-by: Jacek Karwowski <inexxt@users.noreply.github.com>

\begin{definition}[Free group]
  \label{def:presentation}
  Let $A$ be a set, and $\List[\blank]$ the free monoid. The free group $F(A)$ on $A$ is the set-quotient of $\List[A +
      A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons a \cons \nil
    \sim \nil$.
\end{definition}

@@

\begin{definition}[Group presentation]
  Let $A$ be a set and $R : \List[A + A] \to \List[A + A] \to \UU$ a binary relation on $\List[A + A]$. The group
  $F(\langle A ; R \rangle)$ presented by $A$ and $R$, is given by the set-quotient of the free group $F(A)$ by the
  congruence closure of $R$.
\end{definition}

@@

Notice however, that a group presentation, as defined in~\Cref{def:presentation}, requires the relation to be on the set
of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The constructor
$\cancel$ specifies that the inverse of each element is again the same element, using which we can show that our
definition of $\Sn$ is equivalent to the definition of a presented group, by lifting the $\cox*$ relation.

\begin{theorem}
  \leavevmode
  \begin{enumerate}
    \item There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list
          append, and inverse is given by list reversal.
    \item $\Sn$ is equivalent to the group presented by generators $\Fin[n]$ with the relations given by $\cox*$
          extended to $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
  \end{enumerate}
\end{theorem}

To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply have to compute their normal forms using
$\normf$. They correspond to the same permutation if and only if these normal forms are equal, which is decidable for
$\List[\Fin[n]]$.

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
  we compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Lehmer[\zero]          \\
      \eqv & \unit                   \\
      \eqv & \Aut[\Fin[\suc[\zero]]]
    \end{array}
    \qquad\qquad
    \begin{array}{rlr}
           & \Lehmer[\suc[m]]                                                                                    &                                                                 \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                               & \text{by definition}                                            \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                      & \text{induction hypothesis}                                     \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
    \end{array}
  \end{gather*}
\end{proof}

>>>>

\begin{definition}[Free group]
  Let $A$ be a set, and $\List[\blank]$ the free monoid. The free group $F(A)$ on $A$ is the set-quotient of $\List[A +
      A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons a \cons \nil
    \sim \nil$.
\end{definition}

@@

\begin{definition}[Group presentation]
  \label{def:presentation}
  Let $A$ be a set and $R : \List[A + A] \to \List[A + A] \to \UU$ a binary relation on $\List[A + A]$. The group
  $F(\langle A ; R \rangle)$ presented by $A$ and $R$, is given by the set-quotient of the free group $F(A)$ by the
  congruence closure of $R$.
\end{definition}

@@

\noindent
Finally, we need to show that $\Sn$ and is indeed a group presentation.

\begin{propositionrep}
  \label{prop:group-structure-Sn}
  There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list append, and
  inverse is given by list reversal.
\end{propositionrep}

\begin{proof}
   The operation $\append$ is associative, with $\nil$ as the neutral element, and the constructor $\cancel$ makes each
   element an inverse of itself, so the inverse in the group is indeed given by list reversal.
\end{proof}

\noindent
Notice however, that for a group presentation, as defined in~\Cref{def:presentation}, the relation needs to be on the
set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
proposition lets us lift a group structure on $\List[A]$ to a group presentation.

\begin{propositionrep}
  \label{prop:generated-group-generalised}
  Given a set $A$ and a congruence relation $R$ on $\List[A]$ with the group structure given by append and reversal,
  there is a generated group structure $F(\langle A ; \liftrel{R} \rangle)$, where $\liftrel{R}$ is the relation $R$
  lifted to $\List[A+A]$ along the codiagonal map $\nabla_{A} : A + A \to A$.
\end{propositionrep}

\begin{proof}
  Let $G$ be the group on $\List[A]$. We construct the group homomorphism from $F(\langle A ; \liftrel{R} \rangle)$ to
  $G$ using the universal property of the group presentation, where the map on generators is given by taking the
  singleton list. To construct the group homomorphism from $G$, we use the universal property of the set quotient, where
  the map on the underlying set is given by mapping $\inl$ on the list. Using that the relation is a congruence, we show
  that this consttitutes a group isomorphism.
\end{proof}

\begin{corollary}
  The group $\Sn$ is isomorphic to the group presentation given by ${F(\langle \Fin[n] ; \liftrel{\cox*} \rangle)}$.
\end{corollary}

\noindent To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply have to compute their normal forms using
$\normf$. They correspond to the same permutation if and only if these normal forms are equal, which is decidable for
$\List[\Fin[n]]$.

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
  we compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    %% \begin{array}{rl}
    %%        & \Lehmer[\zero]          \\
    %%   \eqv & \unit                   \\
    %%   \eqv & \Aut[\Fin[\suc[\zero]]]
    %% \end{array}
    %% \qquad\qquad
    \begin{array}{rlr}
           & \Lehmer[\suc[m]]                                                                                    &                                                                 \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                               & \text{by definition}                                            \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                      & \text{induction hypothesis}                                     \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
    \end{array}
  \end{gather*}
\end{proof}
